---
title: "PM566 Final Project"
author: "Karisa Ke"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

This is my PM566 Final Project website.

# Introduction
In this project, I'll be analyzing a data set for California housing prices in 1990. This data set contains various features related to housing in different districts in California, such as longitude, latitude, median income, housing median age, median housing value, etc. The questions that I will answer are: 1) is there any correlations between the median income and median housing value; 2) what is the distribution pattern of housing value across California; 3) whether ocean proximity is necessary in this dataset; 4) check if there is correlation between other variables


# Interactive visualizations
```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
library(ggplot2)
library(reshape2)

library(scales) 
```


```{r, message=FALSE, echo=FALSE, warning=FALSE}
housing <- read_csv("data_folder/housing.csv")

haversine <- function(lat1, lon1, lat2, lon2) {
  R <- 6371  # Radius of Earth in km
  phi1 <- lat1 * (pi / 180)
  phi2 <- lat2 * (pi / 180)
  delta_phi <- (lat2 - lat1) * (pi / 180)
  delta_lambda <- (lon2 - lon1) * (pi / 180)
  
  a <- sin(delta_phi / 2)^2 + cos(phi1) * cos(phi2) * sin(delta_lambda / 2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  d <- R * c
  return(d * 1000)  # Convert km to m
}

coordinates <- list(
  'Los Angeles' = c(34.0522, -118.2437),
  'San Diego' = c(32.7157, -117.1611),
  'San Jose' = c(37.3382, -121.8863),
  'San Francisco' = c(37.7749, -122.4194)
)


for (city in names(coordinates)) {
  lat <- coordinates[[city]][1]
  lon <- coordinates[[city]][2]
  housing[[paste('Distance to', city)]] <- mapply(haversine,housing$latitude, housing$longitude, lat, lon)
}

```

## 1
From the correlation matrix, we can see that there is a strong positive correlation between median income and median house value. The higher the income, the more expensive their housing. 

```{r, message=FALSE, echo=FALSE, warning=FALSE}
#  this is for the first  question
numeric_housing <- housing[sapply(housing, is.numeric)]

cor_matrix <- cor(numeric_housing, use = "complete.obs")

melted_cor_matrix <- melt(cor_matrix)


p <- ggplot(melted_cor_matrix, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

p
```

## 2
I extracted the data that falls into the 40%-60% quantile of median_income data. Then I made this scatter plot to show what is the house value of people with medians of median_income. After adding the trend line, it appears that the housing value of people with income between 30k and 40k is 180000-210000. 

```{r, message=FALSE, echo=FALSE, warning=FALSE}
#  this is for the second question


income_40th_percentile <- quantile(housing$median_income, 0.40)
income_60th_percentile <- quantile(housing$median_income, 0.60)


filtered_data <- housing %>%
  filter(median_income >= income_40th_percentile & median_income <= income_60th_percentile)

p <- ggplot(filtered_data, aes(x = median_income, y = median_house_value, color = median_house_value)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red", 
                       breaks = pretty_breaks(n = 5)(range(filtered_data$median_house_value)), 
                       labels = scales::label_comma()) + 
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), color = "black") +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(labels = scales::label_comma()) + # Add this line to format y axis
  theme_minimal()

p

```


```{r, message=FALSE, echo=FALSE, warning=FALSE}
# this is for third question

fig <- plot_ly(data = housing, x = ~longitude, y = ~latitude, z = ~median_house_value, 
               type = 'scatter3d', mode = 'markers',
               marker = list(size = 2, 
                             color = ~median_house_value, 
                             colorscale = 'Bluered',  
                             colorbar = list(title = 'House Price'),
                             opacity = 0.8))

fig <- fig %>% layout(title = '3D Spatial Distribution of House Prices',
                      scene = list(xaxis = list(title = 'Longitude'),
                                   yaxis = list(title = 'Latitude'),
                                   zaxis = list(title = 'Price')))

fig

```

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# this is for fourth question

housing$`Distance to Los Angeles` <- housing$`Distance to Los Angeles` * 0.000621371

housing$`Distance to San Diego` <- housing$`Distance to San Diego` * 0.000621371

housing$`Distance to San Francisco` <- housing$`Distance to San Francisco` * 0.000621371

housing$`Distance to San Jose` <- housing$`Distance to San Jose` * 0.000621371


la_near_housing <- housing %>% 
  filter(`Distance to Los Angeles` <= 200)


p <- ggplot(la_near_housing, aes(x = `Distance to Los Angeles`, y = median_house_value, color = median_house_value)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red",
                       breaks = pretty_breaks(n = 5)(range(la_near_housing$median_house_value)), 
                       labels = label_comma()) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), color = "black") +
  scale_x_continuous(labels = label_comma()) +
  theme_minimal()

p
```



```{r, message=FALSE, echo=FALSE, warning=FALSE}
# for sd 
sd_near_housing <- housing %>% 
  filter(`Distance to San Diego` <= 150)


p_sd <- ggplot(sd_near_housing, aes(x = `Distance to San Diego`, y = median_house_value, color = median_house_value)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red",
                       breaks = pretty_breaks(n = 5)(range(sd_near_housing$median_house_value)), 
                       labels = label_comma()) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), color = "black") +
  scale_x_continuous(labels = label_comma()) +
  theme_minimal()


```


```{r, message=FALSE, echo=FALSE, warning=FALSE}
# this is for fifth question

fig <- plot_ly(data = housing, x = ~longitude, y = ~latitude, z = ~housing_median_age, 
               type = 'scatter3d', mode = 'markers',
               marker = list(size = 2, 
                             color = ~housing_median_age, 
                             colorscale = 'Bluered',  
                             colorbar = list(title = 'House age'),
                             opacity = 0.8))

fig <- fig %>% layout(title = '3D Spatial Distribution of housing_median_age',
                      scene = list(xaxis = list(title = 'Longitude'),
                                   yaxis = list(title = 'Latitude'),
                                   zaxis = list(title = 'housing_median_age')))


fig

p <- ggplot(housing, aes(x = housing_median_age)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Histogram of Housing Median Age",
       x = "Housing Median Age",
       y = "Frequency") +
  theme_minimal()
```

# Download Report
Below is the link to my writtern report: 
[Download PDF](./PM566_Final_Project_ written_report.pdf)
